---
title: "Exam hand in"
author: "Lydwin Wagenaar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)
library(dplyr)
library(glmmTMB)
library(AICcmodavg)
library(MuMIn)
library(car)
library(performance)
library(ggplot2)
library(gridExtra)
```

## Exam Lydwin Wagenaar, Statistics course
Read in the data and remove the variables not necessary for the analysis:
```{r}
Data <- read.csv(file="C:/Users/ly5276wa/Work Folders/Desktop/phd/Courses/Statistics course/Exam/Exam2023_data.csv")
#keep only the variables are of interest
Data.use <- Data[-c(6:7,10:19,21:23,29:38)]
#I want to take into account all the seedlings (an eucalyptus can grow around 1 meter per year and we do not know when they stopped having cattle in the area)
Data.use$tot.Euc <- Data.use$euc_sdlgs0_50cm + Data.use$euc_sdlgs50cm.2m + Data.use$euc_sdlgs.2m

dat.e <- Data.use[-c(11:13)]
```

Explore the data:
```{r}
hist(Data.use$tot.Euc)
#Distribution makes sense because it is count data --> poisson distribution or negative binomial distribution necessary
#There does seem to be an outlier of over 80 seedlings in a plot. We do not have a good reason to remove this outlier yet so we leave it in but need to check how much it affects the results. 

pairs.panels(dat.e[c(6:11)],)
#Aspect and landscape position have a connection which is logical and I will have to choose one of them if I need to include it into the model

#remove rows with NA to have a nice clean dataset
dat.e <- na.omit(dat.e)

#Get some more information about the response variable
range(dat.e$tot.Euc)
mean(dat.e$tot.Euc)
sd(dat.e$tot.Euc)
count(dat.e, tot.Euc == 0)
```
Model selection. The sample size > 40 so we can progress by looking at the AIC:
```{r}
model1 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Aspect + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "poisson")
model2 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Landscape.position + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "poisson")
model3 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Aspect + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "poisson")
model4 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "poisson")
model5 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "poisson")
model6 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Landscape.position + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "poisson")

mlist = list(model1,model2,model3,model4,model5,model6)
aictab(mlist,second.ord = F)
summary(model1)
# Overdispersion problems (deviance >>> residuals) so we will go over to a negative binomial distribution instead
```

Negative binomial model selection:
```{r}
model1 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Aspect + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")
model2 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Landscape.position + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")
model3 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Landscape.position + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")
model4 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")
model5 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")
model6 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Aspect + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")

mlist = list(model1,model2,model3,model4,model5,model6)
aictab(mlist,second.ord = F)
#model 5 and 2 seem pretty good, but 5 seems better than two and is simpler so we will go with this model.

Main.model <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e, na.action = na.exclude, family = "nbinom2")
```

The model results:
```{r}
summary(Main.model)
check_collinearity(Main.model) #no problems with colinearity 
r.squaredGLMM(Main.model) #negative binomial --> can look at lognormal which shows that 14% is explained by fixed effects and 67% is explained by total model.
```
Figure bare ground cover:
```{r}
# Use predict function to create figures based on the model
NewBC1 = seq(min(dat.e$BareGround_cover, na.rm = T), max(dat.e$BareGround_cover, na.rm = T), length.out = length(dat.e$Aspect))
Newdis1 = rep(mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T), length(NewBC1)) #keep the distance from Eu. trees on mean.
NewCC1 = rep(mean(dat.e$Euc_canopy_cover, na.rm = T), length(NewBC1)) #keep the Canopy cover on mean.

Predict = predict(Main.model, newdata=list(BareGround_cover=NewBC1, 
                                Distance_to_Eucalypt_canopy.m.=Newdis1, Euc_canopy_cover=NewCC1, 
                                Season = dat.e$Season, Property = dat.e$Property), 
                type="response",re.form=~0) #we keep the random effects on 0

# now we look at how it is with a mean + sd distance from tree

Newdis.p = rep(mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T) + sd(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm=T), length(NewBC1)) #keep the distance from Eu. trees on mean.

Predict.p = predict(Main.model, newdata=list(BareGround_cover=NewBC1, 
                                Distance_to_Eucalypt_canopy.m.=Newdis.p, Euc_canopy_cover=NewCC1, 
                                Season = dat.e$Season, Property = dat.e$Property), 
                type="response",re.form=~0) #we keep the random effects on 0

# now we look at how it is with a mean - sd distance from tree
Newdis.m = rep(mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T) - sd(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm=T), length(NewBC1))
Predict.m = predict(Main.model, newdata=list(BareGround_cover=NewBC1, 
                                Distance_to_Eucalypt_canopy.m.=Newdis.m, Euc_canopy_cover=NewCC1, 
                                Season = dat.e$Season, Property = dat.e$Property), 
                type="response",re.form=~0) #we keep the random effects on 0


#creating the figure
Colors <- c("Mean" = "black", "Mean + SE" = "darkgoldenrod1", "Mean - SE" = "darkred")
plot.1 <- ggplot(dat.e, aes(x = BareGround_cover, y = tot.Euc)) +
  geom_point(col = "darkgrey") +
  geom_smooth(aes(x = NewBC1, y = Predict,color = "Mean")) +
  geom_smooth(aes(x = NewBC1, y = Predict.p, col = "Mean + SE"), linewidth=0.8, linetype = "dashed") +
  geom_smooth(aes(x = NewBC1, y = Predict.m,col = "Mean - SE"), linewidth=0.8, linetype = "dashed") +
  labs(x = "Bare ground cover (%)", y = "# Eucalyptus seedlings",color = "Distance to tree:") +
    scale_color_manual(values = Colors) + theme(legend.position="none")
plot.1
```
Now for canopy cover:
```{r}
# Use predict function to create figures based on the model
NewCC = seq(min(dat.e$Euc_canopy_cover, na.rm = T), max(dat.e$BareGround_cover, na.rm = T), length.out = length(dat.e$Aspect))
Newdis = rep(mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T), length(NewCC)) #keep the distance from Eu. trees on mean.
NewBC = rep(mean(dat.e$BareGround_cover, na.rm = T), length(NewCC)) #keep the Canopy cover on mean.

Predict.2 = predict(Main.model, newdata=list(BareGround_cover=NewBC, 
                                Distance_to_Eucalypt_canopy.m.=Newdis,
                                Euc_canopy_cover=NewCC, 
                                Season = dat.e$Season, Property = dat.e$Property),
                            type="response",re.form=~0) #we keep the random effects on 0

# now we look at how it is with a mean + sd distance from tree

Newdis.p.2 = rep(mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T) + sd(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm=T), length(NewCC)) #keep the distance from Eu. trees on mean.

Predict.p.2 = predict(Main.model, newdata=list(BareGround_cover=NewBC, 
                                Distance_to_Eucalypt_canopy.m.=Newdis.p.2, 
                                Euc_canopy_cover=NewCC, 
                                Season = dat.e$Season, Property = dat.e$Property), 
                type="response",re.form=~0) #we keep the random effects on 0

# now we look at how it is with a mean - sd distance from tree
Newdis.m.2 = rep(mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T) - sd(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm=T), length(NewCC))
Predict.m.2 = predict(Main.model, newdata=list(BareGround_cover=NewBC, 
                                Distance_to_Eucalypt_canopy.m.=Newdis.m.2, 
                                Euc_canopy_cover=NewCC, 
                                Season = dat.e$Season, Property = dat.e$Property), 
                type="response",re.form=~0) #we keep the random effects on 0

Colors <- c("Mean" = "black", "Mean + SE" = "darkgoldenrod1", "Mean - SE" = "darkred")
plot.2 <- ggplot(dat.e, aes(x = Euc_canopy_cover, y = tot.Euc)) +
  geom_point(col = "darkgrey") +
  geom_smooth(aes(x = NewCC, y = Predict.2,color = "Mean")) +
  geom_smooth(aes(x = NewCC, y = Predict.p.2, col = "Mean + SE"), linewidth=0.8, linetype = "dashed") +
  geom_smooth(aes(x = NewCC, y = Predict.m.2,col = "Mean - SE"), linewidth=0.8, linetype = "dashed") +
  labs(x = "Eucalyptus canopy cover (%)", y = "",color = "Distance to tree") +
    scale_color_manual(values = Colors) + theme(legend.position = c(0.9,0.9))
plot.2

```
End figure:
```{r}
grid.arrange(plot.1, plot.2, ncol=2)

```

What happens if we remove the outlier:
```{r}
dat.e.o <- subset(dat.e, tot.Euc < 80) #removed outlier

model1 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Aspect + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")
model2 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Landscape.position + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")
model3 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Landscape.position + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")
model4 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")
model5 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")
model6 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Aspect + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")

mlist = list(model1,model2,model3,model4,model5,model6)
aictab(mlist,second.ord = F)

# 5 is still the best model so it did not affect model selection too much

Main.model1 <- glmmTMB(tot.Euc ~ Euc_canopy_cover + BareGround_cover + Distance_to_Eucalypt_canopy.m. + (1|Season) + (1|Property), data = dat.e.o, na.action = na.exclude, family = "nbinom2")

summary(Main.model1) # the effects are a bit less strong, but very similar otherwise
#so we will keep the outlier in
r.squaredGLMM(Main.model1) #we can see that the fixed effect do explain less of the model without the outlier which is also what we found in the summary table.

#since we do not have a good reason to expect that this outlier was caused by a measurement error, and it does not change the patterns of the results, we will include the outlier in the analysis.
```

Effect sizes:
```{r}
#Ground cover
coefs <- summary(Main.model)$coef
exp(coefs$cond[1,1] + coefs$cond[3,1]*0 + coefs$cond[2,1]*mean(dat.e$Euc_canopy_cover,na.rm = T) + coefs$cond[4,1]*mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T))
exp(coefs$cond[1,1] + coefs$cond[3,1]*100 + coefs$cond[2,1]*mean(dat.e$Euc_canopy_cover,na.rm = T) + coefs$cond[4,1]*mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T))

#Canopy cover
exp(coefs$cond[1,1] + coefs$cond[2,1]*0 + coefs$cond[3,1]*mean(dat.e$BareGround_cover,na.rm = T) + coefs$cond[4,1]*mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T))
exp(coefs$cond[1,1] + coefs$cond[2,1]*100 + coefs$cond[3,1]*mean(dat.e$BareGround_cover,na.rm = T) + coefs$cond[4,1]*mean(dat.e$Distance_to_Eucalypt_canopy.m.,na.rm = T))

#distance
1-exp(coefs$cond[4,1])
```