---
title: "Mid-term"
author: "Lydwin Wagenaar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gridExtra)
library(AICcmodavg)
library(ggplot2)
library(psych)
library(dplyr)
library(MuMIn)
```

## Mid-term exercise

Read in the data:
```{r}
data <- read.table("C:/Users/ly5276wa/Work Folders/Desktop/phd/Courses/Statistics course/R/CourseStat/tephritis.txt", header = T)
data$Patry <- as.factor(data$Patry)
data$Hostplant <- as.factor(data$Hostplant)
data$Sex <- as.factor(data$Sex)
data$Baltic <- as.factor(data$Baltic)
```

Exploring the data. Since my research question is involving ovipositor length, I will only look at Female flies

```{r}
Female <- subset(data, Sex == "Female")
# 295 observations left
library("psych")
pairs.panels(Female[c(4:5)],)
# Seems to be a connection, I need to include BL in the model as covariate
range(Female$BL)
range(Female$OL, na.rm = T) # There are some NA's present in the OL
Female <- Female[complete.cases(Female[ , c("OL")]), ]
```
Summary statistics
```{r}
mean(Female$OL)
sd(Female$OL)
aggregate(Female$OL, list(Female$Baltic), FUN=mean) 
aggregate(Female$OL, list(Female$Baltic), FUN=sd) 
count(Female.E, Hostplant, Patry)
count(Female.W, Hostplant, Patry) 

```
The model, based on ecological knowledge, would include OL as response variable, and Hostplant, Patry, BL, and Baltic as explanatory variables (Baltic since there are two places where sympatric populations occur). We will do model selection with AIC (N>40) to see which model is fitting the data best and how many parameters we want to include in the model. No random variables are present in this dataset since all categorical variables are < 3 levels. 

```{r}
#Total model just to explore
model <- lm(OL ~ Hostplant * Patry * BL * Baltic, data = Female,na.action = "na.fail")
dredge(model, rank = "AIC")
#chooses a quite complicated model

#Possible models based on knowledge of the ecological system
M1 <- lm(OL ~ Hostplant + Patry + BL + Baltic, data = Female,na.action = "na.fail")
M2 <- lm(OL ~ Hostplant * Patry + BL + Baltic, data = Female,na.action = "na.fail")
M3 <- lm(OL ~ Hostplant * Patry * Baltic + BL, data = Female,na.action = "na.fail")
M4 <- lm(OL ~ Hostplant * Patry * Baltic, data = Female,na.action = "na.fail") 
M5 <- lm(OL ~ Hostplant * Patry + BL, data = Female,na.action = "na.fail")
M6 <- lm(OL ~ Hostplant * Patry, data = Female,na.action = "na.fail") 
M7 <- lm(OL ~ Hostplant + Patry, data = Female,na.action = "na.fail") 
M8 <- lm(OL ~ Hostplant + BL + Baltic, data = Female,na.action = "na.fail")
M9 <- lm(OL ~ Hostplant + BL, data = Female,na.action = "na.fail") 
M10 <- lm(OL ~ Hostplant + Baltic, data = Female,na.action = "na.fail")
M11 <- lm(OL ~ Hostplant + Patry + Baltic, data = Female,na.action = "na.fail")
M12 <- lm(OL ~ 1, data = Female,na.action = "na.fail")
mlist = list(M1, M2, M3, M4, M5, M6, M7, M8, M9, M10, M11, M12)

#perform a AIC test
aictab(mlist,second.ord = F)

```
It seems like the variables that affect ovipositor length depend a lot on where the population occurs. Therefore, we will split the data in two instead.

```{r}
Female.E <- subset(Female, Baltic == "East")
Female.W <- subset(Female, Baltic == "West")

pairs.panels(Female.E[c(4:5)],) #Still normally distributed
pairs.panels(Female.W[c(4:5)],) #Still normally distributed

M1 <- lm(OL ~ Hostplant + Patry + BL, data = Female.E,na.action = "na.fail")
M2 <- lm(OL ~ Hostplant * Patry + BL, data = Female.E,na.action = "na.fail")
M3 <- lm(OL ~ Hostplant * Patry, data = Female.E,na.action = "na.fail") 
M4 <- lm(OL ~ Hostplant + Patry, data = Female.E,na.action = "na.fail") 
M5 <- lm(OL ~ Hostplant + BL, data = Female.E,na.action = "na.fail")
M6 <- lm(OL ~ Hostplant, data = Female.E,na.action = "na.fail")
M7 <- lm(OL ~ 1, data = Female.E,na.action = "na.fail")
mlist = list(M1, M2, M3, M4, M5, M6, M7)

aictab(mlist,second.ord = F)

M1 <- lm(OL ~ Hostplant + Patry + BL, data = Female.W,na.action = "na.fail")
M2 <- lm(OL ~ Hostplant * Patry + BL, data = Female.W,na.action = "na.fail")
M3 <- lm(OL ~ Hostplant * Patry, data = Female.W,na.action = "na.fail") 
M4 <- lm(OL ~ Hostplant + Patry, data = Female.W,na.action = "na.fail") 
M5 <- lm(OL ~ Hostplant + BL, data = Female.W,na.action = "na.fail")
M6 <- lm(OL ~ Hostplant, data = Female.W,na.action = "na.fail")
M7 <- lm(OL ~ 1, data = Female.W,na.action = "na.fail")
mlist = list(M1, M2, M3, M4, M5, M6, M7)

aictab(mlist,second.ord = F)


```
The best model for the Eastern population is:
M5 <- lm(OL ~ Hostplant + BL, data = Female.E,na.action = "na.fail")

The best model for the Western population is:
M2 <- lm(OL ~ Hostplant * Patry + BL, data = Female.W,na.action = "na.fail")

Lets take a closer look at these models.

Western population analysis
```{r}
Model.W <- lm(OL ~ Hostplant * Patry + BL, data = Female.W,na.action = "na.fail")
anova(Model.W)
summary(Model.W)
hist(Model.W$residuals) #pretty normally distributed
# R2 = 59% 
# get some more direct information about the factors their means and SE
Female.W$Patry <- relevel(Female.W$Patry, "Sympatry")
levels(Female.W$Patry)
Female.W$Patry <- relevel(Female.W$Patry, "Allopatry")
try <- lm(OL ~ -1 + Hostplant + Hostplant:Patry + BL, data = Female.W,na.action = "na.fail")
summary(try)
```
Create a figure of the results for the Western population
```{r}
#Predict based on the model
new.data <- data.frame(Patry = "Sympatry", Hostplant = "Heterophyllum", BL = seq(min(Female.W$BL,na.rm = T), max(Female.W$BL,na.rm=T), length.out=20))
newdata1=expand.grid(Patry=c("Sympatry","Allopatry"), Hostplant=c("Heterophyllum","Oleraceum"), BL = seq(min(Female.W$BL,na.rm = T), max(Female.W$BL,na.rm=T), length.out=20))
pre <- predict(Model.W,newdata1,type = "response", se.fit=T)
newdata1$predict <- pre$fit
newdata1$SE <- pre$se.fit

#Create figures
plot.test1 <- ggplot(Female.W, aes(x = BL, y = OL,
           group = Hostplant)) +
           geom_point(aes(col = Hostplant)) + geom_smooth(aes(x = BL, y = predict, color=Hostplant ),data= newdata1, se = F) + geom_smooth(aes(x = BL, y = predict+SE, ),data= newdata1, se = F,linetype="dashed",linewidth=0.5,color="gray60") + geom_smooth(aes(x = BL, y = predict-SE),data= newdata1, se = F, linetype="dashed", linewidth=0.5,color="gray60") +
  scale_color_manual(values=c(Heterophyllum ="darkgoldenrod1",Oleraceum="darkred")) +
  labs(x = "Body length (mm)", y = "Ovipositor length (mm)") +
  ggtitle("West Baltic population") + theme(plot.title = element_text(hjust = 0.5))

Fig.S <- subset(Female.W, Patry == "Sympatry")
Fig.S.T <- subset(newdata1, Patry == "Sympatry")
Fig.A <- subset(Female.W, Patry == "Allopatry")
Fig.A.T <- subset(newdata1, Patry == "Allopatry")

plot.test2 <- ggplot(Fig.S, aes(x = BL, y = OL,
           group = Hostplant)) +
           geom_point(aes(col = Hostplant)) + geom_smooth(aes(x = BL, y = predict, color=Hostplant),data= Fig.S.T, se = F) +  geom_smooth(aes(x = BL, y = predict+SE, ),data= Fig.S.T, se = F,linetype="dashed",linewidth=0.5,color="gray60") + geom_smooth(aes(x = BL, y = predict-SE),data= Fig.S.T, se = F, linetype="dashed", linewidth=0.5,color="gray60") +
  scale_color_manual(values=c(Heterophyllum ="darkgoldenrod1",Oleraceum="darkred")) +
  labs(x = "Body length (mm)", y = "Ovipositor length (mm)") +
  ggtitle("Sympatric population") + theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

plot.test3 <- ggplot(Fig.A, aes(x = BL, y = OL,
           group = Hostplant)) +
           geom_point(aes(col = Hostplant)) + geom_smooth(aes(x = BL, y = predict, color=Hostplant),data= Fig.A.T, se = F) + geom_smooth(aes(x = BL, y = predict+SE, ),data= Fig.A.T, se = F,linetype="dashed",linewidth=0.5,color="gray60") + geom_smooth(aes(x = BL, y = predict-SE),data= Fig.A.T, se = F, linetype="dashed", linewidth=0.5,color="gray60") +
  scale_color_manual(values=c(Heterophyllum ="darkgoldenrod1",Oleraceum="darkred")) +
  labs(x = "Body length (mm)", y = "Ovipositor length (mm)") +
  ggtitle("Allopatric population") + theme(plot.title = element_text(hjust = 0.5),legend.position="none")

grid.arrange(plot.test1, grid.arrange(plot.test2, plot.test3,ncol=2), ncol=1)
  
```
Eastern population analysis

```{r}
Model.E <- lm(OL ~ Hostplant + BL, data = Female.E,na.action = "na.fail")
anova(Model.E)
summary(Model.E)
hist(Model.E$residuals) #looks very good
# get some more direct information about the factors their means and SE
Female.E$Patry <- relevel(Female.E$Patry, "Sympatry")
levels(Female.W$Patry)
Female.E$Patry <- relevel(Female.E$Patry, "Allopatry")
try1 <- lm(OL ~ -1 + Hostplant + BL, data = Female.E,na.action = "na.fail")
summary(try1)
```
Visualize effects of Eastern population
```{r}
#Create predicted data
new.data <- data.frame(Patry = "Sympatry", Hostplant = "Heterophyllum", BL = seq(min(Female.E$BL,na.rm = T), max(Female.E$BL,na.rm=T), length.out=20))
newdata1=expand.grid(Patry=c("Sympatry","Allopatry"), Hostplant=c("Heterophyllum","Oleraceum"), BL = seq(min(Female.E$BL,na.rm = T), max(Female.E$BL,na.rm=T), length.out=20))
pre <- predict(Model.E,newdata1,type = "response", se.fit=T)
newdata1$predict <- pre$fit
newdata1$SE <- pre$se.fit

#Create plots
plot.test1 <- ggplot(Female.E, aes(x = BL, y = OL,
           group = Hostplant)) +
           geom_point(aes(col = Hostplant)) + geom_smooth(aes(x = BL, y = predict, color=Hostplant ),data= newdata1, se = F) + geom_smooth(aes(x = BL, y = predict+SE, ),data= newdata1, se = F,linetype="dashed",linewidth=0.5,color="gray60") + geom_smooth(aes(x = BL, y = predict-SE),data= newdata1, se = F, linetype="dashed", linewidth=0.5,color="gray60") +
  scale_color_manual(values=c(Heterophyllum ="darkgoldenrod1",Oleraceum="darkred")) +
  labs(x = "Body length (mm)", y = "Ovipositor length (mm)") +
  ggtitle("East Baltic population") + theme(plot.title = element_text(hjust = 0.5))

Fig.S <- subset(Female.E, Patry == "Sympatry")
Fig.S.T <- subset(newdata1, Patry == "Sympatry")
Fig.A <- subset(Female.E, Patry == "Allopatry")
Fig.A.T <- subset(newdata1, Patry == "Allopatry")

plot.test2 <- ggplot(Fig.S, aes(x = BL, y = OL,
           group = Hostplant)) +
           geom_point(aes(col = Hostplant)) + geom_smooth(aes(x = BL, y = predict, color=Hostplant),data= Fig.S.T, se = F) +  geom_smooth(aes(x = BL, y = predict+SE, ),data= Fig.S.T, se = F,linetype="dashed",linewidth=0.5,color="gray60") + geom_smooth(aes(x = BL, y = predict-SE),data= Fig.S.T, se = F, linetype="dashed", linewidth=0.5,color="gray60") +
  scale_color_manual(values=c(Heterophyllum ="darkgoldenrod1",Oleraceum="darkred")) +
  labs(x = "Body length (mm)", y = "Ovipositor length (mm)") +
  ggtitle("Sympatric population") + theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

plot.test3 <- ggplot(Fig.A, aes(x = BL, y = OL,
           group = Hostplant)) +
           geom_point(aes(col = Hostplant)) + geom_smooth(aes(x = BL, y = predict, color=Hostplant),data= Fig.A.T, se = F) + geom_smooth(aes(x = BL, y = predict+SE, ),data= Fig.A.T, se = F,linetype="dashed",linewidth=0.5,color="gray60") + geom_smooth(aes(x = BL, y = predict-SE),data= Fig.A.T, se = F, linetype="dashed", linewidth=0.5,color="gray60") +
  scale_color_manual(values=c(Heterophyllum ="darkgoldenrod1",Oleraceum="darkred")) +
  labs(x = "Body length (mm)", y = "Ovipositor length (mm)") +
  ggtitle("Allopatric population") + theme(plot.title = element_text(hjust = 0.5),legend.position="none")

grid.arrange(plot.test1, grid.arrange(plot.test2, plot.test3,ncol=2), ncol=1)
  
```